---
title: "Automating a Pipeline"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`ctoclient` was built for pipelines that run without anyone watching: nightly
data pulls, high-frequency checks, dashboards that refresh on a schedule. This
article covers what changes when nobody is at the console.

## 1. Credentials

An unattended job cannot be prompted, so the password has to come from the
environment. On a server or a scheduled task, set the variables in
`.Renviron`, in the service definition, or in the scheduler:

```{r}
library(ctoclient)

cto_connect(
  server   = Sys.getenv("CTO_SERVER"),
  username = Sys.getenv("CTO_USER"),
  password = Sys.getenv("CTO_PASS")
)
```

Fail early and loudly if they are missing, rather than letting the connection
fail with a less obvious error:

```{r}
required <- c("CTO_SERVER", "CTO_USER", "CTO_PASS")
missing  <- required[!nzchar(Sys.getenv(required))]
if (length(missing)) stop("Missing credentials: ", paste(missing, collapse = ", "))
```

### GitHub Actions

Store each value as a repository secret and expose it to the step:

```yaml
- name: Nightly pull
  env:
    CTO_SERVER: ${{ secrets.CTO_SERVER }}
    CTO_USER:   ${{ secrets.CTO_USER }}
    CTO_PASS:   ${{ secrets.CTO_PASS }}
  run: Rscript scripts/pull.R
```

A private key for an encrypted form does not belong in the repository. Write
it from a secret at run time and delete it afterwards:

```yaml
- name: Restore key
  env:
    CTO_KEY: ${{ secrets.CTO_PRIVATE_KEY }}
  run: |
    printf '%s' "$CTO_KEY" > key.pem
    Rscript scripts/pull.R
    rm -f key.pem
```

Whatever the platform, make sure the account the job uses has only the
permissions it needs. A pipeline that reads submissions does not need rights
to delete datasets.

## 2. Quieting the console

Progress messages are useful interactively and noise in a log:

```{r}
options(ctoclient.verbose = FALSE)
```

Set it once at the top of the script.

## 3. Incremental pulls

`start_date` is applied server-side, so it genuinely reduces the work rather
than filtering after the fact. For a form that grows over months, pulling
everything each night is wasteful and eventually slow.

Keep a watermark and ask only for what is new:

```{r}
state_file <- "state/last_pull.rds"

since <- if (file.exists(state_file)) {
  readRDS(state_file)
} else {
  as.POSIXct("2000-01-01")
}

new_data <- cto_form_data("baseline_survey", start_date = since)

if (nrow(new_data) > 0) {
  saveRDS(max(new_data$SubmissionDate), state_file)
}
```

Two cautions. Write the watermark only *after* the downstream work succeeds,
or a mid-script failure will skip records on the next run. And take the
watermark from `SubmissionDate` rather than the clock, so a submission that
arrives while the job is running is not stepped over.

For anything you will analyse rather than append to, a full pull remains the
safer choice: submissions can be edited or re-approved after the fact, and an
incremental pull will not see those changes.

## 4. Rate limits and parallelism

The session throttles itself to 30 requests per minute. You can raise it, but
check your server tolerates it first:

```{r}
conn <- cto_connect(Sys.getenv("CTO_SERVER"), Sys.getenv("CTO_USER"),
                    Sys.getenv("CTO_PASS"))

conn <- httr2::req_throttle(conn, capacity = 60, fill_time_s = 60)
cto_set_connection(conn)
```

**Do not parallelise across forms.** SurveyCTO rejects concurrent requests
from the same account, so running several pulls at once produces failures
rather than speed. Loop sequentially and let the throttle pace you.

Retries are the useful knob instead:

```{r}
conn <- conn |>
  httr2::req_retry(max_tries = 5) |>
  httr2::req_timeout(300)
cto_set_connection(conn)
```

A long-running form export can take minutes, so raise the timeout before
concluding the server is down.

## 5. Failing well

Each tidying step inside `cto_form_data()` is wrapped individually: if one
fails it prints a message and the rest continue. That is helpful
interactively and dangerous unattended, because a partly-tidied frame is
returned rather than an error.

Check the result before you trust it:

```{r}
data <- cto_form_data("baseline_survey")

stopifnot(
  nrow(data) > 0,
  inherits(data$SubmissionDate, "POSIXct"),
  all(c("hh_id", "resp_age") %in% names(data))
)
```

Asserting the types you depend on turns a silent partial failure into a loud
one. Wrap the whole run so a failure reaches a person:

```{r}
result <- tryCatch(
  {
    run_pipeline()
    "ok"
  },
  error = function(e) {
    notify_team(conditionMessage(e))   # your own code
    "failed"
  }
)
```

## 6. A complete nightly script

```{r}
library(ctoclient)

options(ctoclient.verbose = FALSE)

required <- c("CTO_SERVER", "CTO_USER", "CTO_PASS")
missing  <- required[!nzchar(Sys.getenv(required))]
if (length(missing)) stop("Missing credentials: ", paste(missing, collapse = ", "))

cto_connect(
  server   = Sys.getenv("CTO_SERVER"),
  username = Sys.getenv("CTO_USER"),
  password = Sys.getenv("CTO_PASS")
)

forms <- c("baseline_survey", "followup_survey")

for (form in forms) {
  message("Pulling ", form)

  data <- cto_form_data(form, status = "approved")

  stopifnot(nrow(data) > 0, inherits(data$SubmissionDate, "POSIXct"))

  saveRDS(data, file.path("data", paste0(form, ".rds")))

  cto_form_docx(form, path = file.path("docs", paste0(form, "_review.docx")))
}

message("Done at ", format(Sys.time()))
```

Sequential, quiet, asserted, and with the documentation regenerated from the
deployed definition on every run so it cannot drift.

## 7. Scheduling

On Linux, `cron` calling `Rscript`. On Windows, Task Scheduler. On GitHub
Actions, a `schedule` trigger. In all three, set the working directory
explicitly and make sure `.Renviron` is actually read — a cron job runs with a
minimal environment and often does not pick up the one from your login shell.
Passing the variables in the crontab entry, or sourcing an environment file in
a wrapper script, avoids an evening of confusion.

## See also

- [Managing connections](managing-connections.html) for credential storage
  and multi-server sessions.
- [Working with form data](form-data.html) for what tidying does and where
  it can partially fail.
