---
title: "Documenting and Reviewing a Form"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

A deployed form is a specification, and like any specification it has to be
read by people who did not write it: a field manager checking the wording, a
principal investigator signing off before launch, an analyst working out what
`hh_size_2` meant. `ctoclient` can produce each of those views from the
deployed definition, so what people review is what is actually on the tablets.

```{r}
library(ctoclient)
cto_connect("myorg", "admin@example.com")
```

## 1. The definition itself

Everything here starts from the XLSForm.

```{r}
# Metadata about every deployed and previous version
meta <- cto_form_metadata("baseline_survey")

# The definition file for the version currently deployed
path <- cto_form_definition("baseline_survey")

# Or a specific historical version
old <- cto_form_definition("baseline_survey", version = "20231001")
```

`cto_form_definition()` writes the `.xlsx` to `dir` (the working directory by
default) and returns the path invisibly. It does not re-download a file that
is already there unless you pass `overwrite = TRUE` — convenient in a loop,
surprising if you deployed a new version five minutes ago and wondered why
nothing changed.

Read it with `readxl` if you want the sheets directly:

```{r}
survey   <- readxl::read_excel(path, sheet = "survey")
choices  <- readxl::read_excel(path, sheet = "choices")
settings <- readxl::read_excel(path, sheet = "settings")
```

## 2. A Word document for review

`cto_form_docx()` renders the fields an enumerator actually sees as a
formatted Word table: variable name, label, hint, choice list, relevance and
constraint, with rows shaded by field type and groups and repeats as banded
section headers.

```{r}
cto_form_docx("baseline_survey", path = "baseline_review.docx")
```

Calculates, metadata fields (`start`, `end`, `deviceid`, the audit fields) and
disabled rows are left out, because none of them appear on the tablet. If you
are debugging rather than reviewing and want them back:

```{r}
cto_form_docx("baseline_survey", path = "full.docx", show_metadata = TRUE)
```

The function returns the parsed field table invisibly, which is useful in its
own right — it is a tidy description of the form:

```{r}
fields <- cto_form_docx("baseline_survey", path = tempfile(fileext = ".docx"))
subset(fields, family == "select_one", c(name, label, options))
```

### Changing the colours

Rows are shaded by *family* rather than by raw type, so `select_one yes_no`
and `select_one gender` look the same. `cto_docx_palette()` returns the
defaults:

```{r}
cto_docx_palette()
```

Pass a named vector to override any subset of them; names you leave out keep
their default:

```{r}
cto_form_docx(
  "baseline_survey",
  path = "review.docx",
  palette = c(numeric = "#FFF2CC", note = "#EDEDED")
)
```

### Languages

A multi-language form renders in its `default_language` unless you name one:

```{r}
cto_form_languages("baseline_survey")

cto_form_docx(
  "baseline_survey",
  path = "review_am.docx",
  language = "Amharic (am)"
)
```

## 3. The server's own printable version

SurveyCTO generates its own printable rendering, which `cto_form_printable()`
downloads:

```{r}
cto_form_printable("baseline_survey", relevancies = TRUE, constraints = TRUE)
```

The two are not interchangeable, so it is worth knowing which you want. The
server's version is laid out like the questionnaire a respondent would be
read from, and is the better artefact for field staff and for printing.
`cto_form_docx()` is built locally from the definition and shows variable
names alongside the labels, which is what a reviewer or an analyst needs and
what the server's version does not carry.

## 4. Stata labels

`cto_form_dofile()` writes a `.do` file that labels a dataset exported from
this form: variable labels, value labels, notes, and the loops needed to
reach every indexed copy of a variable inside a repeat group.

```{r}
cto_form_dofile("baseline_survey", path = "baseline_labels.do")
```

It also handles a few things that bite when you do this by hand. Structural
fields that hold no data are dropped, but only after Stata has confirmed every
value is missing. String dates and datetimes are converted, but only if the
variable is still a string, so running the do-file twice cannot wipe them.
Numeric questions are `destring`ed before their value labels are applied.
Labels are escaped so that a `$` or a backtick in a question cannot be
swallowed as a macro reference.

Called without a path, it returns the lines instead of writing them, which is
handy if you want to append to a larger do-file:

```{r}
lines <- cto_form_dofile("baseline_survey")
writeLines(c(readLines("import.do"), lines), "import_labelled.do")
```

For the server's own import template, which sets up reading the CSV in the
first place, use:

```{r}
cto_form_stata_template("baseline_survey")
```

The two complement each other: the template imports, the do-file labels.

## 5. Mail-merge templates

```{r}
cto_form_mail_template("baseline_survey")
```

Useful when a form feeds printed or emailed output and you want the field
placeholders without transcribing them.

## 6. Keeping documentation current

Because every one of these reads the deployed definition, they are cheap to
regenerate and worth regenerating rather than editing by hand. A small script
run after each deployment keeps the review copy honest:

```{r}
for (form in cto_form_ids()) {
  cto_form_docx(form, path = file.path("docs", paste0(form, "_review.docx")))
  cto_form_dofile(form, path = file.path("stata", paste0(form, "_labels.do")))
}
```

`cto_form_ids()` returns the form IDs themselves, so it drops straight into a
loop. For titles, versions and deployment status, use `cto_metadata("forms")`
instead.

## See also

- [Working with form data](form-data.html) for how the same definition
  drives the tidying of submissions.
