---
title: "Managing Server Datasets"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Server datasets are the tables a SurveyCTO server holds alongside your forms:
sampling frames, household rosters, case management lists, pre-loaded choice
lists. `ctoclient` covers the whole lifecycle, and this article walks through
it in the order you would actually meet it.

```{r}
library(ctoclient)
cto_connect("myorg", "admin@example.com")
```

## 1. Seeing what is there

```{r}
datasets <- cto_dataset_list()

info <- cto_dataset_info("sampling_frame")
```

`cto_dataset_list()` returns one row per dataset. `cto_dataset_info()` returns
the configuration of a single one — its title, its unique record field,
whether offline updates are allowed, and so on. Read the second before you
upload to a dataset somebody else set up; the upload modes below behave very
differently depending on how the dataset was configured.

## 2. Creating one

```{r}
cto_dataset_create("sampling_frame")
```

That is the minimum. `title` defaults to the ID, and everything else is
optional — but three arguments are worth knowing about, because they are
awkward to change later:

```{r}
cto_dataset_create(
  "household_cases",
  title               = "Household case management",
  unique_record_field = "hh_id",
  allow_offline_updates = TRUE
)
```

`unique_record_field` names the column that identifies a row. It is what
`MERGE` uploads match on, and what case management uses to find a case. A
dataset created without it can only ever be appended to or replaced wholesale.

`allow_offline_updates` lets enumerators modify records while offline, which
is what you want for case management and usually not what you want for a
static sampling frame.

`discriminator` and the `id_format_options`, `cases_management_options` and
`location_context` lists cover the rest of the server's dataset settings; see
`?cto_dataset_create` for the full set.

## 3. Uploading data

`cto_dataset_upload()` takes a **path to a CSV file**, not a data frame:

```{r}
cto_dataset_upload("sampling_frame", file = "data/frame.csv")
```

To upload something you have in memory, write it out first:

```{r}
tmp <- tempfile(fileext = ".csv")
readr::write_csv(frame, tmp)
cto_dataset_upload("sampling_frame", file = tmp)
```

### The three upload modes

`upload_mode` is the argument to get right, because two of the three can
destroy data.

**`"APPEND"`** (the default) adds the rows in your file to whatever is already
there. It does not check for duplicates. Uploading the same file twice gives
you every row twice.

```{r}
cto_dataset_upload("sampling_frame", "data/new_villages.csv")
```

**`"MERGE"`** matches incoming rows against existing ones and updates them in
place, adding any that do not match. It needs to know what to match on, which
is what `joining_field` is for:

```{r}
cto_dataset_upload(
  "household_cases",
  file          = "data/updated_status.csv",
  upload_mode   = "MERGE",
  joining_field = "hh_id"
)
```

Get `joining_field` wrong and you will either update the wrong rows or append
everything as new records. Check it against `cto_dataset_info()` first.

**`"CLEAR"`** empties the dataset before uploading. Everything currently in it
is gone, and there is no undo:

```{r}
cto_dataset_upload(
  "sampling_frame",
  file        = "data/frame_v2.csv",
  upload_mode = "CLEAR"
)
```

Use it when the file you are uploading is the complete, authoritative
contents. If enumerators have been updating records offline, `CLEAR` discards
their work.

## 4. Downloading

```{r}
# One dataset
cto_dataset_download("sampling_frame", dir = "data")

# Every dataset on the server
cto_dataset_download(dir = "data/backups")
```

Called with no `id`, it downloads all of them — which makes a one-line backup
before a risky upload:

```{r}
cto_dataset_download(dir = "backups/2026-09-19")
cto_dataset_upload("sampling_frame", "data/frame_v2.csv", upload_mode = "CLEAR")
```

Files are written as `<id>.csv`. An existing file is left alone unless you
pass `overwrite = TRUE`, so a scheduled backup will not silently replace
yesterday's copy — and equally, will not refresh it. Use a dated directory as
above, or `overwrite = TRUE` deliberately.

## 5. Emptying and removing

```{r}
# Remove every row, keep the dataset and its configuration
cto_dataset_purge("sampling_frame")

# Remove the dataset itself
cto_dataset_delete("sampling_frame")
```

Both are immediate and neither asks for confirmation. `cto_dataset_purge()`
is the one to reach for when you want to reload a frame but keep the
configuration, the ID and anything referencing it. `cto_dataset_delete()`
breaks any form that pre-loads from that dataset, so check before running it:

```{r}
# What still references this dataset?
cto_metadata("forms")
```

In a script, a purge or a delete is worth guarding behind something explicit
rather than leaving it on a line that can be re-run by accident.

## 6. A worked refresh

Putting it together — a weekly refresh that keeps a backup and replaces the
frame atomically enough to be safe:

```{r}
library(ctoclient)

cto_connect(
  server   = Sys.getenv("CTO_SERVER"),
  username = Sys.getenv("CTO_USER"),
  password = Sys.getenv("CTO_PASS")
)

backup_dir <- file.path("backups", Sys.Date())
dir.create(backup_dir, recursive = TRUE, showWarnings = FALSE)
cto_dataset_download("sampling_frame", dir = backup_dir)

frame <- build_frame()          # your own code
stopifnot(nrow(frame) > 0, !anyDuplicated(frame$hh_id))

tmp <- tempfile(fileext = ".csv")
readr::write_csv(frame, tmp)
cto_dataset_upload("sampling_frame", tmp, upload_mode = "CLEAR")
```

The `stopifnot()` matters more than it looks: `CLEAR` with an empty or
malformed file empties the dataset and puts nothing back.

## See also

- [Automating a pipeline](automation.html) for running this on a schedule.
